### 2014-09-12 Paper_v1.tex
First commit of reduced order coagulation materials

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of Cell-Free Biochemical Networks using Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Joseph A. Wayman, Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective models of metabolism}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\textit{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
Thus, cell-free operation holds several significant advantages for model development, identification and validation. 
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated. 
Toward this unmet need, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
The key innovation of our approach is the integration of simple effective rules encoding complex allosteric regulation with traditional kinetic pathway modeling. 
We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information. 
Second, when integrated into network models, these rules captured classic regulatory patterns such as product-induced feedback inhibition. 
Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.

{\noindent \textbf{Keywords:}~Cell-free metabolism, Mathematical modeling}

@@

\section*{Introduction}
Mathematical modeling has long contributed to our understanding of metabolism. 
Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}. 
The single cell \textit{E. coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}. 
Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E. coli} \citep{2008_atlas_shuler_IETSysBio}.
Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}. 
Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}. 
Unfortunately, cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.

In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}. 
Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E. coli} \citep{Feist:2007aa} or \textit{B. subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}. 
Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.  
Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism. 
Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}. 
Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. 
Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}. However, genome-scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.

Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall. 
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}. 
However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}. 
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}. 
Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system. 
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth. 
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling. 
This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks. 
In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data. 
Toward these questions, we explored five hypothetical cell-free networks. 
Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity. 
We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information. 
Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition. 
Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data. 
While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.

@@

\subsection*{Formulation and properties of effective cell-free metabolic models.}
We developed two proof-of-concept metabolic networks to investigate the features of our effective biochemical network modeling approach (Fig. \ref{fig-networks}). 
In both examples, substrate S was converted to the end products P$_{1}$ and P$_{2}$ through a series of enzymatically catalyzed reactions, including a branch point at hypothetical metabolite M$_{2}$. 
Several of these reactions involved cofactor dependence (AH or A), and various allosteric regulatory mechanisms modified the activity of pathway enzymes. 
Network A included feedback inhibition of the initial pathway enzyme ($E_{1}$) by pathway end products P$_{1}$ and P$_{2}$ (Fig. \ref{fig-networks}A). 
On the other hand, network B involved feedback inhibition of $E_{1}$ by P$_{2}$ and $E_{6}$ by P$_{1}$ (Fig. \ref{fig-networks}B). 
In both networks, branch point enzymes $E_{3}$ and $E_{6}$ were subject to feed-forward activation by reduced cofactor AH. 
Lastly, it is known experimentally that cell-free systems have a finite operational lifespan. 
Loss of biosynthetic capability could be a function of many factors, e.g., cofactor or metabolite limitations. 
We modeled the loss of biosynthetic capability as a non-specific first-order decay of enzyme activity.

Allosteric regulation of enzyme activity was modeled by combining individual regulatory contributions to the activity of pathway enzymes into a control coefficient using an integration rule (Fig. \ref{fig-control-schematic}).  
This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \citep{Morris:2011aa}. 
In our formulation, Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ were used to calculate the influence of factor abundance upon target enzyme activity. 
In this context, factors can be individual metabolite levels or some function, e.g., the product of metabolite levels. 
However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities. 
In the current study, we simply let $\mathcal{Z}$ correspond to the abundance of individual metabolites. 
When an enzyme was potentially sensitive to more than one regulatory input, logical integration rules were used to select which regulatory transfer function influenced enzyme activity at any given time. 
Thus, our test networks involved important features such as cofactor recycling, enzyme activity and metabolite dynamics, as well as multiple overlapping allosteric regulatory mechanisms.

The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction. 
We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.

End product yield was controlled by feedback inhibition, while product selectivity was controlled by branch point enzyme inhibition (Fig. \ref{fig-onoff-simulations}). 
A critical test of our modeling approach was to simulate networks with known behavior. 
If we cannot reproduce the expected behavior of simple networks, then our effective modeling strategy, and particularly the rule-based approximation of allosteric regulation, will not be feasible for genome-scale cell-free problems. We considered two cases, control ON/OFF, for each network configuration. 
Each of these cases had identical kinetic parameters and initial conditions; the \textit{only} differences between the cases were the allosteric regulation rules and the control parameters associated with these rules. 
As expected, end product accumulation was larger for network A when the control was OFF (no feedback inhibition of $E_{1}$ by P$_{1}$ and P$_{2}$),
as compared to the ON case (Fig. \ref{fig-onoff-simulations}A). 
We found this behavior was robust to the choice of underlying kinetic parameters, as we observed that same qualitative response across an ensemble of randomized parameter sets (N = 100), for fixed control parameters. 
The control ON/OFF response of network B was more subtle. 
In the OFF case, the behavior was qualitatively similar to network A. 
However, for the ON case, flux was diverted away from P$_{2}$ formation by feedback inhibition of $E_{6}$ activity at the M$_{2}$ branch point by P$_{1}$ (Fig. \ref{fig-onoff-simulations}B). 
Lower $E_{6}$ activity at the M$_{2}$ branch point allowed more flux toward P$_{1}$ formation, hence the yield of P$_{1}$ also increased (Fig. \ref{fig-onoff-simulations}C). 
Again, the control ON/OFF behavior of network B was robust to changes in kinetic parameters, as the same qualitative trend was conserved across an ensemble of randomized parameters (N = 100), for fixed control parameters. 
Taken together, these simulations suggested that the rule-based allosteric control concept could robustly capture expected feedback behavior for networks with uncertain kinetic parameters.

\subsection*{Estimating parameters and effective allosteric regulatory structures.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
We explored this linkage by estimating model parameters from synthetic data using both network structures. 
We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B). 
However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.

We modified our particle swarm identification strategy to simultaneously search over both kinetic parameters and putative control structures. 
In addition to our initial networks, we constructed three additional presumptive network models, each with the same enzymatic connectivity but different allosteric regulation of the pathway enzymes (Fig. \ref{fig-alternative-networks}). 
We then initialized a population of particles, each with one of the five potential regulatory programs and randomized kinetic parameters. 
Thus, we generated an initial population of particles that had \textit{both} different kinetic parameters as well as different control structures. 
We biased the distribution of the particle population according to our \textit{a prior} belief of the correct regulatory program. 
To this end, we considered three different priors, a uniform distribution where each putative regulatory structure represented 20\% of the population and two mixed distributions that were either positively or negatively biased towards the correct structure (network A). 
In both the positively biased and uniform cases the PSO clearly differentiated between the true or closely related structures and those that were materially different (Fig. \ref{fig-control-search}). 
As expected, the positively biased population (40\% of the initial particle population seeded with network A) gave the best results, where the correct structure was preferentially identified (Fig. \ref{fig-control-search}A). 
On the other hand, when given a uniform distribution, the PSO approach identified a combination of network A and network C as the most likely control structures (Fig. \ref{fig-control-search}B). 
Network A and C differ by the regulatory connection between the end product P$_2$ and enzyme $E_{1}$; in network A, end product P$_{2}$ was assumed to inhibit $E_{1}$, while in network C, end product P$_{2}$ activated $E_{1}$.  
Lastly, when the initial population was biased towards incorrect structures (initial population seeded with 90\% incorrect structures), the particle swarm \textit{misidentified} the correct allosteric structure (Fig. \ref{fig-control-search}C). 
Interestingly, while each particle swarm identified parameter sets that minimized the simulation error, the estimated parameter values were not necessarily similar to the true parameters. 
The angle between the estimated and true parameters was not consistently small across the swarms (identical parameters would give an angle of zero). 
This suggested that our particle swarm approach identified a \textit{sloppy} ensemble, i.e., parameter estimates that were individually incorrect but collectively exhibited the correct model behavior.

We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig. \ref{fig-flux-pattern}C). 
In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control. 
Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity. 
Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate. Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network. 
On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure. 
This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Networks.pdf}
\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products P$_{1}$ and P$_{2}$ through a series of chemical conversions
catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-6-AlternativeNetworks.pdf}
\caption{Schematic of the alternative allosteric control programs used in the structural particle swarm computation. 
Each network had the same enzymatic connectivity, initial conditions and kinetic parameters, 
but alternative feedback control structures for the first enzyme in the pathway.}\label{fig-alternative-networks}
\end{figure}

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort&compress,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

{\noindent \textbf{Keywords:}~Coagulation, Mathematical modeling, Systems Biology}

@@

%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
%Many have suggested that the integration of experimental and computational research is required to unravel
%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
%like thrombosis during surgery.
%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
%Development and validation of detailed network models requires a significant investment of time and material resources. 
%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.

\section*{Introduction}
The human coagulation system is an archetype proteolytic cascade regulated by both positive and negative feedback loops.
The biology of coagulation is well studied \cite{ROBERTS1998,MANN1999,BUTENAS2002,FURIE2004}. 
Coagulation is mediated by a family serine proteases, called factors and a key group of blood cells, called platelets. 
The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
Trigger events such as vessel injury expose procoagulant materials like collagen, fibronectin, von Willebrand factor (vWF) or tissue factor (TF).
These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
Once coagulation has been activated, two converging pathways activate the master protease thrombin. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Thrombin generation consists of three phases, initiation, propagation and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
Initially thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
Picomolar amounts of thrombin can then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
resulting in the formation of the \emph{tenase} and \emph{prothrombinase} complexes on the surface of activated platelets.
These complexes amplify the early coagulation signal by activating downstream and upstream coagulation factors.  
Termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

Malfunctions in coagulation can have serious or potentially even fatal consequences.
For example, aggressive clotting is involved with Coronary Artery Diseases (CADs), which collectively account for 38\% of all deaths in North America \cite{HANSSON2005}.
Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
However, clotting factor replacement sometimes leads to the formation of \emph{in-vivo} fVIII and fIX inhibitors \cite{TOMOKIYO2003}.
Activated Prothrombin Complex Concentrates (aPCCs) from pooled plasma have been used to treat hemophilia. 
However, aPCCs can have safety issues \cite{CHAVIN1988,ALLEN2006,SCHNEIDERMAN2007}. 
Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
hemophilia with and without factor VIII/IX inhibitors \cite{Lusher:1998bs,Shapiro:1998sh,Shapiro:2000pr,Shapiro:2008px}.
However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
Many questions also remain about its mechanism \cite{KOMIYAMA1990,PAYNE1996,MONROE1997,HOFFMAN1998,BUTENAS2002a,ROBERTS2004,ROBERTS2006,ALLEN2007}
and effective dose range \cite{TOMOKIYO2003}.
Thus, despite its wide adoption, the utility of rFVIIa in trauma-associated hemorrhage remains controversial \cite{Duchesne:2008hw}.

Modeling approaches differ in their degree of mechanistic detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
The ability of ODE modeling to capture dynamics has made it one of the most common tools for studying signaling systems. 
However, dynamics and ODEs come at the expense of difficult (and often impossible) to solve model parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
This reality highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
Moreover, detailed ODE models require significant mechanistic knowledge of the underlying biology. 
However, as Bailey suggested more than a decade ago, achieving qualitative and quantitative understanding of complex biological systems should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability were sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable model predictions could be possible with limited parameter information. 
Taking advantage of this property of model behavior, we developed sloppy techniques for parameter identification using ensembles of deterministic models. 
Pareto optimal ensemble techniques (POETs) incorporate principles of competing objectives into a multi-objective optimization framework, enabling the exploration of kinetic parameter space while accounting for uncertainty and potential conflicts in the experimental training data \citep{2010_song_varner_BiotechJ}. 
We have proposed that the sloppy behavior of biological networks may also be a source of cell-to-cell \citep{2011_lequieu_chakrabarti_nayak_varner_PLoSCompBio} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}. 
Recently, Bayesian techniques of parameter identification have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo}. 
Thus, a population of cells could be viewed as a dynamic ensemble of networks as the operational biochemical pathways are often context-specific \citep{2012_creixell_linding_NatBiotech}.

In this study, we constructed a reduced order effective model of thrombin generation in normal and hemophilic blood.
We used this model to understand how physiological coagulation was altered by prothrombin levels, the activated protein C pathway and the level of ATIII.  
Previous models of coagulation have been formulated as systems of nonlinear ordinary differential equations, often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE]. 
While this approach has proven highly effective, it does require significant experimental data for model identification and validation.    
For example, a previous coagulation model from our laboratory, which described 193 proteins and protein complexes interconnected by 301 biochemical interactions, 
contained 467 unknown parameters (301 kinetic parameters and 166 initial conditions) [REFHERE]. 
The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulatory motifs with traditional kinetic pathway modeling. 
This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information. 
The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions. 
We tested our approach by modeling the time evolution of the human coagulation cascade.

@@

\subsection*{Formulation and properties of reduced order effective models.}
Reduced order effective models are [FINISH ME].

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
\caption{Proof of concept cell-free metabolic networks considered in this study. Substrate $S$ is converted to products $P_{1}$ and $P_{2}$ through a series of chemical conversions
catalyzed by enzyme(s) $E_{j}$. The activity of the pathway enzymes is subject to both positive and negative allosteric regulation. }\label{fig-networks}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-6-Hemophilia.pdf}
\caption{Reduced order coagulation model predictions versus experimental data with and without FVIII and FIX.
The reduced order coagulation model parameter estimates were tested against data not used during model training. 
Simulations of normal thrombin formation with ATIII and the protein C pathway were compared with thrombin formation in the absence of fVIII and fIX.
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
(A,B) prediction results for normal thrombin generation and thrombin generation in hemophilia. 
All factors and control proteins were at their physiological concentration unless others noted.
The experimental validation data was reproduced from the study of Allen et al. \citep{ALLEN2006}.}\label{fig-hemophilia}
\end{figure}
